Estimating dark matter halo concentration using the integrated mass profile

### 2015-11-11 paper/paper.tex
puliendo

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title{Estimating dark matter halo concentration using the integrated
  mass profile} 
\author[C. Poveda, J.E. Forero-Romero, J.C. Mu\~noz-Cuartas]{
\parbox[t]{\textwidth}{\raggedright 
  C. Poveda \thanks{cn.poveda542@uniandes.edu.co}$^{1}$
  J. E. Forero-Romero \thanks{je.forero@uniandes.edu.co}$^{1}$
  J. C. Mu\~noz-Cuartas \thanks{xxxx@udea.edu.co}$^{2}$
}
\vspace*{6pt}\\
$^1$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\
$^2$Instituto de F\'{\i}sica - FCEN, Universidad de Antioquia, Calle
67 No. 53-108, Medell\'{\i}n, Colombia
}

@@

\begin{abstract}

We present a new algorithm to estimate the concentration of N-body
dark matter halos based on the radially integrated mass profile
Its main advantage is that it uses the full particle information
without any binning, making it realiable in  cases when low numerical
resolution becomes a limitation for other methods.  Additionally, the
algorithm is straightforward to implement.  
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution, outperforming the other
two methods. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. 
For the maximum radial velocity method it is on the order of $20\%$
while density profile methods fail at the $100\%$ level.
For halos with $2\times 10^4$ particles our method achieves an
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods 
At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

In this paper we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This results in a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters, providing a new independent
method to estimate the concentration parameter. 
Here we provide a detailed description of this method and a
comparison against the two traditional methods by measuring its
performance both on mock and N-body halos.

@@

In this paper we use $\Delta_h=740$, a number roughly corresponding to
200 times the critical density at redshift z=0.

@@

We now review the key assumptions of the different methods used
to estimate the concentration before presenting our proposal.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations. 
The first method tries to directly estimate the density profile.  
It takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.  
At this point is possible to make a direct fit to the density as a
function of the radial coordinate. 
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

@@

Our method uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).
To build it from N-body data we follow the next steps.
First, we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.  
Then we rank the particles by their increasing radial distance from
the center. 
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. In
this process we discard the particle at the center.

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter. 
We use an Affine Invariant Markov Chain Monte Carlo implemented in the
python module {\em emcee} \citep{emcee} to sample the likelihood
function distribution defined by ${\cal L}(c)\propto
\exp(-\chi^2(c)/2)$ where the $\chi^2(c)$ is 
written as

@@

The construction of the numerical integrated mass profile has two
advantages with respect to the density and velocity methods, it does
not involve any binning (unlike the density methods) and the sampling
method uses the information from all the particles in the halo (unlike
the velocity method).

@@

We now present the results of applying the three methods above
mentione on two different halo samples.  The first sample is composed
by mock halos generated to have known concentration values in perfect
spherical symmetry following an NFW profile.  We use this sample to
check that we can recover the expected values but also gauge the
impact of the number of particles on the outcomes.  The second sample
comes from a publicly available N-body cosmological simulation.  From
this sample we quantify again the differences between all the methods
and also estimate the impact on the mass-concentration relationship.

@@

In this paper, we presented a new method to estimate the concentration
of dark matter halos in N-body simulations.  We tested our method on
mock halo data to study the impact of total number of particles and
input concentration on the retrieved values.  We compared these
results against two other methods commonly used in the literature to
estimate concentrations.  Finally, we applied our method to halos
extracted from a cosmological N-body simulation to estimate the impact
of our method on the mass concentration relationship.

>>>>

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title[Halo concentration from the mass profile]{Estimating dark matter
  halo concentration using the integrated mass profile} 
\author[C. Poveda, J.E. Forero-Romero, J.C. Mu\~noz-Cuartas]{
\parbox[t]{\textwidth}{\raggedright 
  C. Poveda \thanks{cn.poveda542@uniandes.edu.co}$^{1}$
  J. E. Forero-Romero \thanks{je.forero@uniandes.edu.co}$^{1}$
  J. C. Mu\~noz-Cuartas \thanks{xxxx@udea.edu.co}$^{2}$
}
\vspace*{6pt}\\
$^1$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\
$^2$Instituto de F\'{\i}sica - FCEN, Universidad de Antioquia, Calle
67 No. 53-108, Medell\'{\i}n, Colombia
}

@@

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos.
The method uses the integrated mass profile to find the optimal
concentration value. 
Its main advantage is that it uses the full particle information
without any binning, making it realiable in  cases when low numerical
resolution becomes a limitation for other methods.  
Additionally, the algorithm is straightforward to implement.  
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution, outperforming the other
two methods. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy, while for the maximum radial
velocity method it is on the order of $20\%$ 
and density profile methods fail at the $100\%$ level.
For halos with $2\times 10^4$ particles our method achieves a recovery
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We also measure the mass-concentration relationship on N-body
data. We find that within the probed mass range ($10^{12}\hMsun <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
We only find a small discrepancy at low masses, $M<10^{13}\hMsun$, where
our method yields lower median concentration values by $15\%-20\%$
compared to the velocity and density methods. 
From these results we believe that the new method deserves to be
included in a simulator's toolkit to probe the internal structure of
dark matter halos.
\end{abstract}

@@

In this letter we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This results in a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters, providing a new independent
method to estimate the concentration parameter. 
Here we provide a detailed description of this method and a
comparison against the two traditional methods by measuring its
performance both on mock and N-body halos.

@@

In this letter we use $\Delta_h=740$, a number roughly corresponding to
200 times the critical density at redshift z=0.

@@

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.

The first method tries to directly estimate the density profile.  
It takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.  
At this point is possible to make a direct fit to the density as a
function of the radial coordinate. 
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

@@

Our method uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).
We build it from N-body data following the next steps.
First, we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.  
Then we rank the particles by their increasing radial distance from
the center. 
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. 
In this process we discard the particle at the center.

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter. 
We then use an Affine Invariant Markov Chain Monte Carlo implemented in the
python module {\em emcee} \citep{emcee} to sample the likelihood
function distribution defined by ${\cal L}(c)\propto
\exp(-\chi^2(c)/2)$ where the $\chi^2(c)$ is 
written as

@@

@@

We apply the three methods mentioned above on two different halo
samples.
The first sample is composed by mock halos generated to have known
concentration values in perfect spherical symmetry following an NFW
profile.    
Using this sample we want to quantify the ability to recover the
expected concentration as a function of particle numbe.
The second sample comes from a publicly available N-body cosmological
simulation.  
In this way we test the new method on more realistic and complex
data. 
We summarize the results on this dataset by showing the impact on the
mass-concentration relationship.

@@

In this letter we presented a new method to estimate the concentration
of dark matter halos in N-body simulations.  
We tested our method on mock halo data to study the impact of total
number of particles and input concentration on the retrieved values.  
We compared these
results against two other methods commonly used in the literature to
estimate concentrations.  Finally, we applied our method to halos
extracted from a cosmological N-body simulation to estimate the impact
of our method on the mass concentration relationship.
